---
title: "05_KM_curves_log_rank_tests"
author: "Trio"
date: "2024-05-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(survival)
library(survminer)
library(dplyr)
```

```{r}
METABRIC_SUBSET$RFS_STATUS <- as.numeric(METABRIC_SUBSET$RFS_STATUS)
```

KM curve based on receptor subtype and overall survival (p= 9e-14). From the curve, we can see
lower survival rates for HER2+ and LumB. This observation aligns with the absence of Trastuzumab 
in the therapeutic regimens for patients testing positive for HER2

```{r}
fit_KM_by_receptor <- survfit(Surv(overall_survival_years, overall_survival) ~ 
                                receptor_subtype, data= METABRIC_SUBSET,
                              subset = receptor_subtype %in% c("HER2+", "LumA", "LumB", "TNBC"))
ggsurvplot(fit_KM_by_receptor, risk.table.col = "strata", surv.median.line = "hv", conf.int = FALSE) 
fit_log_rank_receptor <- survdiff(Surv(overall_survival_years, overall_survival) ~ 
                           receptor_subtype, data= METABRIC_SUBSET,
                         subset =  receptor_subtype %in% c("HER2+", "LumA", "LumB", "TNBC"))
fit_log_rank_receptor 
```
KM curve based on receptor subtype and RFS (p= 1e-08). The pattern is very similar
to the previous one


```{r}
fit_KM_by_RFS_receptor <- survfit(Surv(RFS_years, RFS_STATUS) ~ 
                                receptor_subtype, data= METABRIC_SUBSET,
                              subset = receptor_subtype %in% c("HER2+", "LumA", "LumB", "TNBC"))

ggsurvplot(fit_KM_by_RFS_receptor, risk.table.col = "strata", surv.median.line = "hv", conf.int = FALSE) 

fit_log_rank_RFS_receptor <- survdiff(Surv(RFS_years, RFS_STATUS) ~ 
                           receptor_subtype, data= METABRIC_SUBSET,
                         subset =  receptor_subtype %in% c("HER2+", "LumA", "LumB", "TNBC"))

fit_log_rank_RFS_receptor
```

KM curve based on neoplasm histologic grade and overall survival (p= 8e-10). 
The test is performed considering all the patients. This difference is surely related to
the distribution of the grades among the receptor subtypes (see the chi-squared test previously performed)

```{r}
fit_KM_by_grade <- survfit(Surv(overall_survival_years, overall_survival) ~ 
                                    neoplasm_histologic_grade, data= METABRIC_SUBSET)
ggsurvplot(fit_KM_by_grade, risk.table.col = "strata", surv.median.line = "hv", conf.int = FALSE) 

fit_log_rank_grade <- survdiff(Surv(overall_survival_years, overall_survival) ~ 
                                        neoplasm_histologic_grade, data= METABRIC_SUBSET)

fit_log_rank_grade
```

KM curve based on neoplasm histologic grade and RFS (p= 2e-06). The pattern is very similar
to the previous one

```{r}
fit_KM_by_RFS_grade <- survfit(Surv(RFS_years, RFS_STATUS) ~ 
                             neoplasm_histologic_grade, data= METABRIC_SUBSET)
ggsurvplot(fit_KM_by_RFS_grade, risk.table.col = "strata", surv.median.line = "hv", conf.int = FALSE) 
fit_log_rank_RFS_grade <- survdiff(Surv(RFS_years, RFS_STATUS) ~ 
                                 neoplasm_histologic_grade, data= METABRIC_SUBSET)
fit_log_rank_RFS_grade
```

The same tests were performed also for cellularity, but in that case we do not have 
evidence to reject H0

Laterality is non significant, as we already know from the previous from semester.
Same destiny for the histological subtypes (we never assessed them, so I tried just to be sure)




Despite the number of NA, I also tried for tumor stage, considering its importance
from a prognostic point of view. We have 975 patients out of 1285 with an available tumor stage
(1 = 320, 2 = 559, 3= 88, 4 = 8). We can ask to Alessia or Cappozzo if we can proceed
but I don't think there will be any problems (maybe if we want to introduce itin the Cox model). 

KM curve based on tumor stage and overall survival (p= <2e-16). It was clearly expected this result

```{r}
fit_KM_by_stage <- survfit(Surv(overall_survival_years, overall_survival) ~ 
                             tumor_stage, data= METABRIC_SUBSET,
                           subset = tumor_stage %in% c("1", "2", "3", "4"))
ggsurvplot(fit_KM_by_stage, risk.table.col = "strata", surv.median.line = "hv", conf.int = FALSE) 
fit_log_rank_stage <- survdiff(Surv(overall_survival_years, overall_survival) ~ 
                                 tumor_stage, data= METABRIC_SUBSET,
                               subset = tumor_stage %in% c("1", "2", "3", "4"))
fit_log_rank_stage
```
KM curve based on tumor stage and RFS (p= <2e-16)

```{r}
fit_KM_by_RFS_stage <- survfit(Surv(RFS_years, RFS_STATUS) ~ 
                             tumor_stage, data= METABRIC_SUBSET,
                           subset = tumor_stage %in% c("1", "2", "3", "4"))
ggsurvplot(fit_KM_by_RFS_stage, risk.table.col = "strata", surv.median.line = "hv", conf.int = FALSE) 
fit_log_rank_RFS_stage <- survdiff(Surv(RFS_years, RFS_STATUS) ~ 
                                 tumor_stage, data= METABRIC_SUBSET,
                               subset = tumor_stage %in% c("1", "2", "3", "4"))
fit_log_rank_RFS_stage

```

Considering the results obtained regarding tumor stage, we can surely stratify 
within tumor size and pos lymph nodes. Besides, we can also evaluate a KM curve based on NPI


KM curve based on tumor size and overall survival (p= <2e-16)
```{r}
fit_KM_by_size <- survfit(Surv(overall_survival_years, overall_survival) ~ 
                             tumor_size_stage, data= METABRIC_SUBSET)
ggsurvplot(fit_KM_by_size, risk.table.col = "strata", surv.median.line = "hv", conf.int = FALSE) 
fit_log_rank_size <- survdiff(Surv(overall_survival_years, overall_survival) ~ 
                                tumor_size_stage, data= METABRIC_SUBSET)
fit_log_rank_size
```
KM curve based on tumor size and RFS (p= 1e-13)

```{r}
fit_KM_by_RFS_size <- survfit(Surv(RFS_years, RFS_STATUS) ~ 
                            tumor_size_stage, data= METABRIC_SUBSET)
ggsurvplot(fit_KM_by_RFS_size, risk.table.col = "strata", surv.median.line = "hv", conf.int = FALSE) 
fit_log_rank_RFS_size <- survdiff(Surv(RFS_years, RFS_STATUS) ~ 
                                    tumor_size_stage, data= METABRIC_SUBSET)
fit_log_rank_RFS_size
```
KM curve based on pos lymph nodes and overall survival (p= <2e-16)

```{r}
fit_KM_by_pos_nodes <- survfit(Surv(overall_survival_years, overall_survival) ~ 
                            pos_lymph_nodes_stage, data= METABRIC_SUBSET)
ggsurvplot(fit_KM_by_pos_nodes, risk.table.col = "strata", surv.median.line = "hv", conf.int = FALSE) 
fit_log_rank_pos_nodes <- survdiff(Surv(overall_survival_years, overall_survival) ~ 
                                     pos_lymph_nodes_stage, data= METABRIC_SUBSET)
fit_log_rank_pos_nodes
```

KM curve based on pos lymph nodes and RFS (p= <2e-16)

```{r}
fit_KM_by_RFS_pos_nodes <- survfit(Surv(RFS_years, RFS_STATUS) ~ 
                                 pos_lymph_nodes_stage, data= METABRIC_SUBSET)
ggsurvplot(fit_KM_by_RFS_pos_nodes, risk.table.col = "strata", surv.median.line = "hv", conf.int = FALSE) 
fit_log_rank_RFS_pos_nodes <- survdiff(Surv(RFS_years, RFS_STATUS) ~ 
                                         pos_lymph_nodes_stage, data= METABRIC_SUBSET)
fit_log_rank_RFS_pos_nodes
```

KM curve based on surgery treatment and overall survival (p= 6e-11). The pattern aligns with the clinical insights: mastectomy is performed on patients with more spread and aggressive breast cancer (lower probability to survive). Is it worth enough to put in the final work? We will see it

```{r}
fit_KM_by_surgery <- survfit(Surv(overall_survival_years, overall_survival) ~ 
                                type_of_breast_surgery, data= METABRIC_SUBSET,
                             subset = type_of_breast_surgery %in% c("MASTECTOMY", "BREAST CONSERVING"))
ggsurvplot(fit_KM_by_surgery, risk.table.col = "strata", surv.median.line = "hv", conf.int = FALSE) 
fit_log_rank_surgery <- survdiff(Surv(overall_survival_years, overall_survival) ~ 
                                   type_of_breast_surgery, data= METABRIC_SUBSET,
                                 subset = type_of_breast_surgery %in% c("MASTECTOMY", "BREAST CONSERVING"))
fit_log_rank_surgery
```
KM curve based on surgery treatment and RFS (p= 4e-08)

```{r}
fit_KM_by_RFS_surgery <- survfit(Surv(RFS_years, RFS_STATUS) ~ 
                               type_of_breast_surgery, data= METABRIC_SUBSET,
                             subset = type_of_breast_surgery %in% c("MASTECTOMY", "BREAST CONSERVING"))
ggsurvplot(fit_KM_by_RFS_surgery, risk.table.col = "strata", surv.median.line = "hv", conf.int = FALSE) 
fit_log_rank_RFS_surgery <- survdiff(Surv(RFS_years, RFS_STATUS) ~ 
                                       type_of_breast_surgery, data= METABRIC_SUBSET,
                                     subset = type_of_breast_surgery %in% c("MASTECTOMY", "BREAST CONSERVING"))
fit_log_rank_RFS_surgery
```
KM curve based on NPI categorized and overall survival (p= <2e-16)

```{r}
fit_KM_by_NPI <- survfit(Surv(overall_survival_years, overall_survival) ~ 
                               NPI_stage, data= METABRIC_SUBSET)
ggsurvplot(fit_KM_by_NPI, risk.table.col = "strata", surv.median.line = "hv", conf.int = FALSE) 
fit_log_rank_NPI <- survdiff(Surv(overall_survival_years, overall_survival) ~ 
                               NPI_stage, data= METABRIC_SUBSET)
fit_log_rank_NPI
```











