---
title: "Logistic Regression"
author: "Elena Sasso, Luca Lepore, Marco Cominelli"
date: "2024-06-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(pROC)
library(caret)
library(glmnet)
library(car)
```

Dataset preparation 

```{r}
# Creation of the variable "relapse_within_5yr"
METABRIC_SUBSET$relapse_within_5yr <- with(METABRIC_SUBSET, ifelse(overall_survival_years <= 5 & overall_survival == 1, 1, ifelse(overall_survival_years > 5, 0, NA)))

# Creation of the relapse dataframe and removal of patients with RFS status = 0 and RFS years < 5 years
METABRIC_SUBSET_relapse <- METABRIC_SUBSET[,c(1:39, 215, 40:214)]

# Make dichotomous the gene mutations values
for(i in 47:215) {
  METABRIC_SUBSET_relapse[[i]] <- ifelse(METABRIC_SUBSET_relapse[[i]] != "0", "1", "0")
}

# Removal of the non meaningful variables
METABRIC_SUBSET_relapse <- METABRIC_SUBSET_relapse[,-c(1,4,5,9,11,13,15,24,25,28,30,31,33,34,35,36)]

# Removal of NA values
METABRIC_SUBSET_relapse <- METABRIC_SUBSET_relapse[!is.na(METABRIC_SUBSET$relapse_within_5yr) & METABRIC_SUBSET$primary_tumor_laterality!="" & METABRIC_SUBSET$type_of_breast_surgery!="",]
METABRIC_SUBSET_relapse <- na.omit(METABRIC_SUBSET_relapse)

```


Make as factor our variables and merge some categories in some of them

```{r}
METABRIC_SUBSET_relapse$cellularity <- factor(METABRIC_SUBSET_relapse$cellularity, levels = c("Low", "Moderate", "High"))
METABRIC_SUBSET_relapse$type_of_breast_surgery <- as.factor(METABRIC_SUBSET_relapse$type_of_breast_surgery)
METABRIC_SUBSET_relapse$chemotherapy <- as.factor(METABRIC_SUBSET_relapse$chemotherapy)
METABRIC_SUBSET_relapse$er_status_measured_by_ihc <- as.factor(METABRIC_SUBSET_relapse$er_status_measured_by_ihc)
METABRIC_SUBSET_relapse$neoplasm_histologic_grade <- factor(METABRIC_SUBSET_relapse$neoplasm_histologic_grade, levels = c("1", "2", "3"))
METABRIC_SUBSET_relapse$her2_status <- as.factor(METABRIC_SUBSET_relapse$her2_status)
METABRIC_SUBSET_relapse$hormone_therapy <- as.factor(METABRIC_SUBSET_relapse$hormone_therapy)
METABRIC_SUBSET_relapse$inferred_menopausal_state <- as.factor(METABRIC_SUBSET_relapse$inferred_menopausal_state)
METABRIC_SUBSET_relapse$integrative_cluster <- as.factor(METABRIC_SUBSET_relapse$integrative_cluster)
METABRIC_SUBSET_relapse$primary_tumor_laterality <- as.factor(METABRIC_SUBSET_relapse$primary_tumor_laterality)
METABRIC_SUBSET_relapse$pr_status <- as.factor(METABRIC_SUBSET_relapse$pr_status)
METABRIC_SUBSET_relapse$radio_therapy <- as.factor(METABRIC_SUBSET_relapse$radio_therapy)
METABRIC_SUBSET_relapse$pik3ca_mut <- as.factor(METABRIC_SUBSET_relapse$pik3ca_mut)
METABRIC_SUBSET_relapse$tp53_mut <- as.factor(METABRIC_SUBSET_relapse$tp53_mut)
METABRIC_SUBSET_relapse$brca_mut <- as.factor(METABRIC_SUBSET_relapse$brca_mut)
METABRIC_SUBSET_relapse$akt_mut <- as.factor(METABRIC_SUBSET_relapse$akt_mut)
METABRIC_SUBSET_relapse$pten_mut <- as.factor(METABRIC_SUBSET_relapse$pten_mut)
METABRIC_SUBSET_relapse$cdh1_mut <- as.factor(METABRIC_SUBSET_relapse$cdh1_mut)

for(i in 31:199) {
  METABRIC_SUBSET_relapse[[i]] <- as.factor(METABRIC_SUBSET_relapse[[i]])
}

cols_to_remove <- sapply(METABRIC_SUBSET_relapse, function(x) is.factor(x) && length(levels(x)) < 2)
METABRIC_SUBSET_relapse <- METABRIC_SUBSET_relapse[, !cols_to_remove]


METABRIC_SUBSET_relapse$pam50_._claudin.low_subtype <- as.factor(METABRIC_SUBSET_relapse$pam50_._claudin.low_subtype)
METABRIC_SUBSET_relapse$pam50_._claudin.low_subtype <- relevel(METABRIC_SUBSET_relapse$pam50_._claudin.low_subtype, ref = "LumA")

METABRIC_SUBSET_relapse$NPI_stage <- as.character(METABRIC_SUBSET_relapse$NPI_stage)
METABRIC_SUBSET_relapse$NPI_stage[METABRIC_SUBSET_relapse$NPI_stage %in% c("Excellent", "Good")] <- "Excellent_Good"
METABRIC_SUBSET_relapse$NPI_stage <- factor(METABRIC_SUBSET_relapse$NPI_stage, levels = c("Excellent_Good", "Moderate", "Poor"))
METABRIC_SUBSET_relapse$NPI_stage <- relevel(METABRIC_SUBSET_relapse$NPI_stage, ref = "Excellent_Good")

METABRIC_SUBSET_relapse$tumor_size_stage <- as.character(METABRIC_SUBSET_relapse$tumor_size_stage)
METABRIC_SUBSET_relapse$tumor_size_stage[METABRIC_SUBSET_relapse$tumor_size_stage %in% c("T2", "T3")] <- "T2_T3"
METABRIC_SUBSET_relapse$tumor_size_stage <- factor(METABRIC_SUBSET_relapse$tumor_size_stage, levels = c("T1", "T2_T3"))

METABRIC_SUBSET_relapse$neoplasm_histologic_grade <- as.character(METABRIC_SUBSET_relapse$neoplasm_histologic_grade)
METABRIC_SUBSET_relapse$neoplasm_histologic_grade[METABRIC_SUBSET_relapse$neoplasm_histologic_grade %in% c("1", "2")] <- "1_2"
METABRIC_SUBSET_relapse$neoplasm_histologic_grade <- factor(METABRIC_SUBSET_relapse$neoplasm_histologic_grade, levels = c("1_2", "3"))

METABRIC_SUBSET_relapse$pos_lymph_nodes_stage <- as.character(METABRIC_SUBSET_relapse$pos_lymph_nodes_stage)
METABRIC_SUBSET_relapse$pos_lymph_nodes_stage[METABRIC_SUBSET_relapse$pos_lymph_nodes_stage %in% c("pN2", "pN3")] <- "pN2_pN3"
METABRIC_SUBSET_relapse$pos_lymph_nodes_stage <- factor(METABRIC_SUBSET_relapse$pos_lymph_nodes_stage, levels = c("pN0", "pN1", "pN2_pN3"))

METABRIC_SUBSET_relapse$pam50_._claudin.low_subtype <- as.character(METABRIC_SUBSET_relapse$pam50_._claudin.low_subtype)
METABRIC_SUBSET_relapse$pam50_._claudin.low_subtype[METABRIC_SUBSET_relapse$pam50_._claudin.low_subtype %in% c("Basal", "Her2")] <- "Basal_Her2"
METABRIC_SUBSET_relapse$pam50_._claudin.low_subtype[METABRIC_SUBSET_relapse$pam50_._claudin.low_subtype %in% c("claudin-low", "Normal")] <- "claudin-low_Normal"
METABRIC_SUBSET_relapse$pam50_._claudin.low_subtype <- factor(METABRIC_SUBSET_relapse$pam50_._claudin.low_subtype)
METABRIC_SUBSET_relapse$pam50_._claudin.low_subtype <- relevel(METABRIC_SUBSET_relapse$pam50_._claudin.low_subtype, ref = "LumA")
```


Univariate models to select just the meaningful variables

```{r}
covariates <- colnames(METABRIC_SUBSET_relapse)[!colnames(METABRIC_SUBSET_relapse) %in% c("relapse_within_5yr")]
models <- list()
pvalues <- list()

for (covariate in covariates) {
  # glm model
  formula <- as.formula(paste("relapse_within_5yr ~", covariate))
  glm_model <- glm(formula, family=binomial(link=logit), data = METABRIC_SUBSET_relapse)
  
  # save the model
  models[[covariate]] <- glm_model
  
  # take the p-value
  pvalue <- summary(glm_model)$coefficients[2,4]
  
  # save the p-value p-value
  pvalues[[covariate]] <- pvalue
}


test <- p.adjust(pvalues, method = "BH")
p_selection <- ifelse(test <= 0.05, test, NA)
p_significant <- p_selection[!is.na(p_selection)]
significant_variables <- names(p_significant)


selected_variables <- c(significant_variables, "relapse_within_5yr")
METABRIC_SUBSET_new_relapse <- METABRIC_SUBSET_relapse[, selected_variables, drop = FALSE]
```

Working on the training set

```{r}
set.seed(345)

# Index for the training and test groups
train_index <- createDataPartition(METABRIC_SUBSET_new_relapse$relapse_within_5yr, p = 0.7, list = FALSE, times = 1)

# Creation of the training and test groups
train_set <- METABRIC_SUBSET_new_relapse[train_index, ]
test_set <- METABRIC_SUBSET_new_relapse[-train_index, ]



# Creation of the full logistic model on the training set just to evaluate the collinearity
glm_model <- glm(relapse_within_5yr ~ . , family = binomial(link = logit), data = train_set)

# Compute VIF
print(vif(glm_model))

# Removal of the collinear variables
train_set <- train_set[,-c(3,8,9,10,11,14)]
test_set <- test_set[,-c(3,8,9,10,11,14)]

# Convert in matrix format
x_train <- model.matrix(relapse_within_5yr ~ ., train_set)[,-1] # rimuovi l'intercept
y_train <- train_set$relapse_within_5yr

# Optimal lambda value through cross validation
cv.out <- cv.glmnet(x_train, y_train, alpha = 1, family = "binomial", type.measure = "mse")
lambda_min <- cv.out$lambda.min
cat("Lambda min:", lambda_min, "\n")

# Coefficients 
print(coef(cv.out, s = lambda_min))


# Check if variables are significant
logistic_final <- glm(relapse_within_5yr ~ er_status_measured_by_ihc +
        neoplasm_histologic_grade + her2_status + tumor_size_stage + pos_lymph_nodes_stage +
        tp53_mut, family=binomial(link=logit), data = METABRIC_SUBSET_new_relapse)
summary(logistic_final)

# Prediction on the training set and conversion in vector
train_prob <- predict(cv.out, newx = x_train, s = lambda_min, type = "response")
train_prob <- as.vector(train_prob)

# Compute ROC curve and AUC on the training set
roc_curve_train <- roc(y_train, train_prob)
auc_train <- auc(roc_curve_train)
print(paste("AUC:", auc_train))
plot(roc_curve_train, main = "ROC Curve for Lasso Logistic Regression on Test Set")
ci(roc_curve_train)

# Select the best threshold. We select it in order to obtain 85% as sensitivity because we want a screening test that is able to reduce at most the false negatives without impact too much the specificity. Diagnostic second level test can be then used to avoid false positives
best_coords_train <- coords(roc_curve_train, x = 0.137, transpose = TRUE)
best_threshold <- best_coords_train["threshold"]

```

Working on the test set

```{r}
# Conversion in matrix format for the test set
x_test <- model.matrix(relapse_within_5yr ~ ., test_set)[,-1] # rimuovi l'intercept
y_test <- test_set$relapse_within_5yr

# Prediction on test set and conversion in vector
lasso_prob <- predict(cv.out, newx = x_test, s = lambda_min, type = "response")
lasso_prob <- as.vector(lasso_prob)

# Apply the selected threshold
lasso_predict <- ifelse(lasso_prob > best_threshold, "1", "0")

# Confusion matrix
conf_matrix <- table(pred = lasso_predict, true = y_test)
print(conf_matrix)

# Compute the ROC curve and AUC
roc_curve_test <- roc(y_test, lasso_prob)
auc_test <- auc(roc_curve_test)
cat("AUC sul test set:", auc_test, "\n")

# Plot
plot(roc_curve_test, main = "ROC Curve for Lasso Logistic Regression on Test Set")
ci(roc_curve_test)
```

Considering the high number of selected variables, we try to eliminate some of them based on the clinical significance without impact too much our model. So, we perform the same procedure again.

```{r}
set.seed(345)

# Index for the training and test groups
train_index <- createDataPartition(METABRIC_SUBSET_new_relapse$relapse_within_5yr, p = 0.7, list = FALSE, times = 1)

# Creation of the training and test groups
train_set <- METABRIC_SUBSET_new_relapse[train_index, ]
test_set <- METABRIC_SUBSET_new_relapse[-train_index, ]

# Removal of the less significant variables from a clinical point of view + the ones from multicollinearity
train_set <- train_set[,-c(1,2, 3, 7,8,9,10,11, 14, 16, 17, 18, 19)]
test_set <- test_set[,-c(1, 2, 3, 7,8,9,10,11, 14, 16, 17, 18, 19)]

# Conversion in matrix format
x_train <- model.matrix(relapse_within_5yr ~ ., train_set)[,-1] # rimuovi l'intercept
y_train <- train_set$relapse_within_5yr

# Optimal lambda through cross validation
cv.out <- cv.glmnet(x_train, y_train, alpha = 1, family = "binomial", type.measure = "mse")

lambda_min <- cv.out$lambda.min
cat("Lambda min:", lambda_min, "\n")

# Coefficients
print(coef(cv.out, s = lambda_min))

# Prediction on the training set
train_prob <- predict(cv.out, newx = x_train, s = lambda_min, type = "response")
train_prob <- as.vector(train_prob)

# Compute ROC curve and AUC
roc_curve_train <- roc(y_train, train_prob)
auc_train <- auc(roc_curve_train)
print(paste("AUC:", auc_train))
plot(roc_curve_train, main = "ROC Curve for Lasso Logistic Regression on Test Set")
ci(roc_curve_train)

# Choosing the threshold as before, so I want a 85% of sensitivity
best_coords_train <- coords(roc_curve_train, x = 0.15, transpose = TRUE)
best_coords_train
best_threshold <- best_coords_train["threshold"]

# Conversion in matrix format
x_test <- model.matrix(relapse_within_5yr ~ ., test_set)[,-1] # rimuovi l'intercept
y_test <- test_set$relapse_within_5yr

# Prediction on test set
lasso_prob <- predict(cv.out, newx = x_test, s = lambda_min, type = "response")
lasso_prob <- as.vector(lasso_prob)

# Apply the selected threshold
lasso_predict <- ifelse(lasso_prob > best_threshold, "1", "0")

# Confusion matrix
conf_matrix <- table(pred = lasso_predict, true = y_test)
print(conf_matrix)

# Compute ROC curve and AUC
roc_curve_test <- roc(y_test, lasso_prob)
auc_test <- auc(roc_curve_test)
cat("AUC sul test set:", auc_test, "\n")

# Plot
plot(roc_curve_test, main = "ROC Curve for Lasso Logistic Regression on Test Set")
ci(roc_curve_test)
```

